---
title: "ADA_Lab1"
author: "Alaina Flory"
date: "2025-08-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

```

## 1

Import class one survey data from our Github site (https://github.com/kijohnson/Advanced-Data-Analysis) (The dataset is called ‘Class 1 Survey Fall 2025 2.csv’), calling the R dataframe that you create C1survey.

```{r}

C1survey <- read.csv("https://github.com/kijohnson/Advanced-Data-Analysis/raw/refs/heads/main/Class%201%20Survey%20Fall%202025%202.csv")

```

## 2

Determine: a. the number of observations (i.e. the number of people who filled out the survey) and b. the number of variables in the dataframe.

```{r}
##number of observations and variable

dim(C1survey)

## 58 observations, 27 variables

```

## 3
a. Rename the column variables to something shorter and that is descriptive of what the variable is about (for example like_dogs for the ‘Do you like dogs?’ question variable) and b. Display that they are renamed. Hints: You can use either the rename_with function or the names function. The list of names for time is:
“id”, “like_cats”, “like_dogs”, “have_desert”, “slogan”, “fav_day”, “larkORowl”, “fav_food”, “fav_drink”, “fav_season”, “fav_month”, “hobby”, “program”, “specialization”, “stat_software”, “R_exp”, “coding_comfort”, “coding_length”, “top_three”, “public_health_interest”, “fav_num”, “bday”, “bmonth”, “country”, “state”, “city”, “highest_educ_level”

```{r}

##renaming columns with new names

names(C1survey)[1:27] <- c("id", "like_cats", "like_dogs", "have_desert", "slogan", "fav_day", "larkORowl", "fav_food", "fav_drink", "fav_season", "fav_month", "hobby", "program", "specialization", "stat_software", "R_exp", "coding_comfort", "coding_length", "top_three", "public_health_interest", "fav_num", "bday", "bmonth", "country", "state", "city", "highest_educ_level")

```

## 4

Write code to determine and display the number of factor, integer, numerical, and character variables there are in the C1survey dataset. Suggestion: Try to ask ChatGPT for a solution if you are unsure. There is a really short answer to this but we want to see what students come up with!

```{r}

#Create a table to display the number of each variable type

table(sapply(C1survey, class))

```

5. a. Using code, check the bday and bmonth variables for any unusual or missing values (hint: use table function). If you find any, b. describe what you will do with the unusual values in a comment before or immediately following the code that does it, and c. after you have cleaned up any unusual values, find the median bday and month. *Note: please make sure that your bday and month variables are numeric before doing statistics

```{r}

#check bday variable values, correct format should be number only
table(C1survey$bday)

# Extract the first number in the string (the day) to fix format and change to numeric
C1survey <- C1survey %>%
  mutate(bday = as.numeric(str_extract(bday, "\\d+")))

#check bmonth variable values
table(C1survey$bmonth)

#Change the months that are spelled out into their corresponding number and change to numeric
C1survey <- C1survey %>%
  mutate(bmonth = case_when(
    bmonth == "January"  ~ 1,
    bmonth == "February" ~ 2,
    bmonth == "March"    ~ 3,
    bmonth == "April"    ~ 4,
    bmonth == "May"      ~ 5,
    bmonth == "June"     ~ 6,
    bmonth == "june"     ~ 6,
    bmonth == "July"     ~ 7,
    bmonth == "August"   ~ 8,
    bmonth == "Septemeber"~ 9,
    bmonth == "October"  ~ 10,
    bmonth == "November" ~ 11,
    bmonth == "December" ~ 12,
    TRUE ~ as.numeric(bmonth) 
  ))

#check variable class and make sure it is changed to numeric
class(C1survey$bday)
class(C1survey$bmonth)


#Find median values of bday and bmonth
median(C1survey$bday, na.rm = TRUE)
median(C1survey$bmonth, na.rm = TRUE)

```

6. 
a. Create a new variable called bseason that gives the season according to Northern Meteorological season in which respondents were born (winter = 12/1 to 2/29, spring = 3/1 to 5/31, summer = 6/1 to 8/31, fall = 9/1 to 11/30). 
b. Using the table function, print a table of Seasons in the columns and bmonths in the rows that allows you to check that the coding is correct 
c. Sum the columns using the addmargins function to answer the question of how many classmates were born in each season? Include your answer in the code.

```{r}

#create bseason variable 

C1survey <- C1survey %>%
   mutate(
    bseason = case_when(
      (bmonth == 12 | bmonth <= 2) ~ "Winter",
      (bmonth >= 3 & bmonth <= 5) ~ "Spring",
      (bmonth >= 6 & bmonth <= 8) ~ "Summer",
      (bmonth >= 9 & bmonth <= 11) ~ "Fall",
      TRUE ~ NA_character_  # in case bmonth is NA
    )
  )

#print table of seasons and months

seasons_table <- table(C1survey$bmonth, C1survey$bseason)

print(seasons_table)

#Find number of people born in each season
addmargins(seasons_table)

##Number of people in class born per season
#Fall = 15
#Spring = 16
#Summer = 16
#Winter = 10

```

7. Pick your favorite variable or variables to analyze, come up with a question you want to answer with that variable or variables, generate the code, and provide an answer to your question. Describe what you did using comments (i.e. #’s).

```{r}

##Question: Does level of comfort in their coding ability correlate with how long they have been coding?

#Create a table with coding_length as the columns and coding_comfort as the rows
coding_table <- table(C1survey$coding_comfort, C1survey$coding_length)
print(coding_table)

#Add sums to the table to see the totals of how many people per coding_length and coding_comfort category
addmargins(coding_table)

##Results: Most people in the class (24 people) have little to no experience in R coding, and the majority of those with little experience responded with low comfort with R coding as well (coding_comfort = 1 or 2).

```

